---
title: "Project 2"
author: "Maria Halai"
date: "2024-11-08"
output: html_document
---
```{r}
library(jsonlite)
library(dplyr)
library(ggplot2)

path <- "/Users/mariahalai/Desktop/gps/"
files <- list.files(path, pattern = "*.geojson", 
                    full.names = TRUE)

data_frames <- list()
for (i in seq_along(files)) {
  file <- files[i]
  dat <- fromJSON(file)
  records <- data.frame(
    file = basename(file),
    day_label = paste("Day", i), 
    time = dat$features$properties$time,
    time_long = dat$features$properties$time_long,
    provider = dat$features$properties$provider,
    accuracy = dat$features$properties$accuracy,
    altitude = dat$features$properties$altitude,
    bearing = dat$features$properties$bearing,
    speed = dat$features$properties$speed,
    longitude = sapply(dat$features$geometry$coordinates, 
                       function(coord) coord[1]),
    latitude = sapply(dat$features$geometry$coordinates, 
                      function(coord) coord[2])
  )
  data_frames[[i]] <- records
}

gps_data <- bind_rows(data_frames)
gps_data$time <- as.POSIXct(gps_data$time, 
                            format="%Y-%m-%dT%H:%M:%OSZ", 
                            tz="UTC")

gps_data <- gps_data %>%
  mutate(day_label = factor(day_label, 
                            levels = paste("Day", 1:11)))

head(gps_data)
```

1.1. Visualize the GPS Path (Latitude vs. Longitude)
This plot will show the routes taken each day based on GPS coordinates.

```{r}
library(ggplot2)
library(dplyr)

ggplot(gps_data, aes(x = longitude, 
                     y = latitude, 
                     color = day_label)) +
  geom_point(alpha = 0.5, 
             size = 1) +
  labs(title = "GPS Paths Over Two Weeks by Day", 
       x = "Longitude", 
       y = "Latitude", 
       color = "Day") +
  theme_minimal()
```
1.2. Analyze Speed Over Time

```{r}
ggplot(gps_data, aes(x = time, 
                     y = speed, 
                     color = day_label)) +
  geom_line(alpha = 0.7) +
  geom_point(data = subset(gps_data, 
                           speed == 0), 
             aes(x = time, 
                 y = speed), 
             color = "red", 
             size = 1.5, 
             alpha = 0.6) +
  labs(title = "Speed Over Time with Stationary Periods Highlighted", 
       x = "Time", 
       y = "Speed (m/s)", color = "Day") +
  theme_minimal() +
  theme(legend.position = "right")
```
1.3. Identify Stationary Periods and Calculate Movement Duration and Transit Periods

1. Identify Movement and Stationary Data:
2. Find Stationary Periods of 2+ Minutes:

```{r}
library(jsonlite)
library(dplyr)
library(tidyr)

gps_data <- gps_data %>%
  mutate(status = ifelse(is.na(speed) | speed == 0, "Stationary", "Moving"))

time_summary <- gps_data %>%
  group_by(day_label, status) %>%
  summarize(total_time = n()) %>%
  pivot_wider(names_from = status, values_from = total_time, 
              values_fill = 0) %>%
  rename(total_movement_time = Moving, 
         total_stationary_time = Stationary)

stationary_periods <- gps_data %>%
  arrange(day_label, time) %>%  
  group_by(day_label) %>%
  mutate(
    stationary_period = cumsum(ifelse(status == "Stationary" & lag(status, 
                                                                   default = "Moving") == "Moving", 1, 0))
  ) %>%
  group_by(day_label, stationary_period) %>%
  filter(status == "Stationary") %>%
  summarize(stationary_duration = n()) %>%
  filter(stationary_duration >= 10) %>%
  ungroup() %>%
  select(day_label, stationary_period, 
         stationary_duration)

print(time_summary)
print(stationary_periods)
```


1.4. Visualize Transit and Stationary Times

```{r}
ggplot(time_summary, aes(x = day_label)) +
  geom_bar(aes(y = total_movement_time, 
               fill = "Transit"), 
           stat = "identity", 
           alpha = 0.6) +
  geom_bar(aes(y = total_stationary_time, 
               fill = "Stationary"), 
           stat = "identity", 
           alpha = 0.6) +
  labs(title = "Transit vs. Stationary Time Per Day", 
       x = "Day", y = "Time Points") +
  scale_fill_manual(values = c("Transit" = "blue", "Stationary" = "red"), 
                    name = "Time Type") +
  theme_minimal()
```
Step 1: Analyze Start and End Times for Transit Periods

Goal: Determine if there’s a consistent daily pattern for when the target starts moving and when transit typically ends.

```{r}
transit_summary <- gps_data %>%
  filter(status == "Moving") %>%
  group_by(day_label) %>%
  summarize(
    avg_start_time = min(time),
    avg_end_time = max(time)
  )
print("Transit Start and End Times by Day:")
print(transit_summary)
```

Step 2: Examine the Relationship Between Speed and Transit Duration

Goal: Determine if longer transit periods correspond to higher speeds, which may indicate faster routes or different transit modes

```{r}
transit_speed_summary <- gps_data %>%
  filter(status == "Moving") %>%
  group_by(day_label) %>%
  summarize(
    avg_speed = mean(speed, na.rm = TRUE),
    transit_duration = n()
  )

ggplot(transit_speed_summary, 
       aes(x = transit_duration, y = avg_speed)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship Between Transit Duration and Average Speed",
       x = "Transit Duration (Time Points)",
       y = "Average Speed (m/s)") +
  theme_minimal()

```

Step 3: Identify Stationary “Safe Zones”
Goal: Pinpoint locations where stationary periods frequently occur, creating “safe zones” to avoid bombing attempts

```{r}
library(dplyr)

gps_data <- gps_data %>%
  arrange(file, time) %>%
  group_by(file) %>%
  mutate(
    stationary_period = ifelse(status == "Stationary" & 
                                 (lag(status, default = "Moving") != "Stationary"), 
                               cumsum(status == "Stationary"), NA)
  ) %>%
  fill(stationary_period, .direction = "down") 

stationary_periods <- gps_data %>%
  filter(status == "Stationary" & !is.na(stationary_period)) %>%
  group_by(file, stationary_period) %>%
  summarize(stationary_duration = n(), .groups = "drop") %>%
  filter(stationary_duration >= 10)  

gps_data <- gps_data %>%
  left_join(stationary_periods, by = c("file", "stationary_period"))

stationary_locations <- gps_data %>%
  filter(status == "Stationary" & !is.na(stationary_duration)) %>%
  group_by(latitude, longitude) %>%
  summarize(stationary_count = n(), .groups = "drop") %>%
  arrange(desc(stationary_count))

head(stationary_locations)
```

```{r}
transit_locations <- gps_data %>%
  filter(status == "Moving") %>%
  group_by(longitude, latitude) %>%
  summarize(transit_count = n()) %>%
  arrange(desc(transit_count)) %>%
  ungroup()

print(head(transit_locations))

library(stats)

coordinates <- transit_locations %>% select(longitude, latitude)
set.seed(123) 
k_clusters <- 5 
clusters <- kmeans(coordinates, 
                   centers = k_clusters)

transit_locations$cluster <- clusters$cluster
```

TEMPERATURE DATA ADDED

```{r}
gps_data$date <- as.Date(gps_data$time)  

unique_dates <- unique(gps_data$date)

print("Dates with available GPS data:")
print(unique_dates)
```

Adding in Temperature data
```{r}
temperature_data <- data.frame(
  date = as.Date(c("2020-08-18", "2020-08-19", "2020-08-20", 
                   "2020-08-21", "2020-08-22", "2020-08-24", 
                   "2020-08-25", "2020-08-26", "2020-08-27", 
                   "2020-08-28", "2020-08-31")),
  max_temp = c(91, 91, 82, 94, 85, 83, 75, 87, 87, 86, 58),
  min_temp = c(65, 61, 57, 49, 56, 53, 50, 49, 52, 46, 46),
  avg_temp = c(78, 76, 69.5, 71.5, 70.5, 68, 62.5, 68, 69.5, 66, 52)
)

gps_data$date <- as.Date(gps_data$time)

gps_data <- gps_data %>%
  left_join(temperature_data, 
            by = "date")

head(gps_data)
```
TEMPERATURE VS SPEED
```{r}
speed_temp_analysis <- gps_data %>%
  group_by(date) %>%
  summarize(avg_speed = mean(speed, na.rm = TRUE), 
            avg_temp = first(avg_temp)) %>%
  ungroup()

library(ggplot2)
ggplot(speed_temp_analysis, aes(x = avg_temp, 
                                y = avg_speed)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, 
              color = "blue") +
  labs(title = "Average Speed vs. Average Temperature",
       x = "Average Temperature (°F)",
       y = "Average Speed (m/s)") +
  theme_minimal()

correlation <- cor(speed_temp_analysis$avg_speed, 
                   speed_temp_analysis$avg_temp)
print(paste("Correlation between speed and temperature:", 
            round(correlation, 2)))

model <- lm(avg_speed ~ avg_temp, 
            data = speed_temp_analysis)
summary(model)
```

```{r}
gps_data <- gps_data %>%
  mutate(temp_category = case_when(
    avg_temp < 60 ~ "Cold",
    avg_temp >= 60 & avg_temp < 75 ~ "Moderate",
    avg_temp >= 75 ~ "Warm"
  ))

anova_result <- aov(speed ~ temp_category, data = gps_data)
summary(anova_result)
```

```{r}
transit_times <- gps_data %>%
  filter(status == "Moving") %>%
  group_by(date) %>%
  summarize(start_time = min(time), 
            end_time = max(time)) %>%
  ungroup()

print("Transit times for each day:")
print(transit_times)

```
Step 1: Calculate the Most Frequent Start Times for Movement
Use the GPS data from the first two weeks to find the most frequent or average start time for movement. This would be the earliest time of day that the target begins moving.
```{r}
movement_start_summary <- gps_data %>%
  filter(speed > 0) %>%
  group_by(date) %>%
  summarize(start_time = min(time)) %>%
  ungroup() %>%
  summarize(frequent_start_time = mean(start_time))

print(movement_start_summary)
```

Step 2: Identify Frequent Transit Locations
Identify locations where the target is typically in transit (not stationary for long periods). We can analyze the coordinates recorded during movement and calculate the most common transit points.
```{r}
transit_locations <- gps_data %>%
  filter(speed > 0) %>%
  group_by(longitude, latitude) %>%
  summarize(transit_count = n()) %>%
  arrange(desc(transit_count)) %>%
  top_n(5, transit_count)

head(transit_locations)
```
Step 3: Calculate Average Transit Duration
Find the typical duration of the target’s movement each day. This will help us determine how long the target might stay in transit, allowing us to target mid-transit times for the bombing attempt.
```{r}
movement_duration_summary <- gps_data %>%
  filter(speed > 0) %>%
  group_by(date) %>%
  summarize(movement_duration = max(time) - min(time)) %>%
  summarize(avg_movement_duration = mean(movement_duration))

print(movement_duration_summary)
```
CALCULATING DISTANCE 
```{r}
if (!requireNamespace("fields", quietly = TRUE)) {
  install.packages("fields")
}
library(fields)

calculate_distance <- function(predicted_location, 
                               actual_location) {
  dist <- rdist.earth(
    matrix(predicted_location, ncol = 2), 
    matrix(actual_location, ncol = 2),
    miles = FALSE  
  )
  return(dist * 1000)  
}
```
FINAL ALGORITHM!!!!! --- Try your attempts here 

```{r}
# These libraries are required for data manipulation, date handling, and geographic calculations
library(dplyr)        # For data manipulation
library(lubridate)    # For date and time handling
library(geosphere)    # For geographic calculations (distance and bearing)

# Function to calculate the bomb time based on average transit duration
# This ensures the bomb is placed at mid-transit while avoiding the first 5 minutes of movement
calculate_bomb_time <- function(start_time, avg_duration) {
  mid_transit_time <- start_time + (avg_duration / 2)  # Mid-transit time based on average duration
  bomb_time <- mid_transit_time + 300  # Offset by 5 minutes to respect the rule on initial movement period
  return(bomb_time)
}

# Function to predict the bombing location based on common transit points
# Uses the centroid of common transit points to approximate the most frequent path
predict_bomb_location <- function(start_location, common_points) {
  # If there are multiple common points, calculate the centroid as the bomb location
  if (nrow(common_points) > 0) {
    centroid <- colMeans(common_points[, c("longitude", "latitude")])
  } else {
    # If no common points are provided, default to start location
    centroid <- start_location
  }
  return(centroid)
}

# Function to check if the bomb location is near any stationary (safe) zone
# Returns TRUE if the location is within 5 meters of a stationary zone, otherwise FALSE
is_near_stationary <- function(bomb_location, stationary_locations, buffer_distance = 5) {
  # Loop through each stationary location and calculate the distance to the bomb location
  for (i in 1:nrow(stationary_locations)) {
    stationary_point <- stationary_locations[i, c("longitude", "latitude")]
    distance_to_stationary <- distVincentySphere(stationary_point, bomb_location)
    
    # If the distance is within the buffer, return TRUE indicating it's too close
    if (distance_to_stationary <= buffer_distance) {
      return(TRUE)  # Bomb location is near a stationary location
    }
  }
  return(FALSE)  # Bomb location is not near any stationary location
}

# Main function to generate the bomb plan based on start location, start time, and other parameters
# This function outputs the day of the week, time of explosion, longitude, and latitude for bombing
generate_bomb_plan <- function(start_location, start_time, avg_duration, buffer_distance, time_buffer, stationary_locations, common_points) {
  # Step 1: Calculate the bomb time based on average transit duration
  bomb_time <- calculate_bomb_time(start_time, avg_duration)
  
  # Step 2: Predict the bomb location using frequent transit points
  bomb_location <- predict_bomb_location(start_location, common_points)
  
  # Check if bomb_location was successfully created
  if (is.null(bomb_location) || length(bomb_location) != 2) {
    stop("Error: 'bomb_location' not properly defined.")
  }
  
  # Step 3: Check distance constraint and avoid stationary zones
  # Calculate the distance from the start location to the predicted bomb location
  distance_to_target <- distVincentySphere(start_location, bomb_location)
  
  # If the bomb location is either too far or too close to a stationary zone, adjust it
  if (distance_to_target > buffer_distance) {
    message("Adjusting bomb location to ensure it's within 5 meters of the transit path.")
    # Adjust location by moving bomb_location within 5 meters from start_location if needed
    bomb_location <- geosphere::destPoint(start_location, bearing(start_location, bomb_location), buffer_distance)
  }
  
  # Step 4: Ensure the bomb location is not in a stationary (safe) zone
  if (is_near_stationary(bomb_location, stationary_locations, buffer_distance)) {
    message("Bomb location is within a safe zone. Adjusting location to avoid stationary areas.")
    bomb_location <- geosphere::destPoint(start_location, bearing(start_location, bomb_location), buffer_distance)
  }
  
  # Step 5: Confirm bomb time falls within transit period and adjust if necessary
  if (bomb_time < start_time + minutes(5) || bomb_time > start_time + avg_duration) {
    message("Adjusting bomb time to be within the active transit period.")
    bomb_time <- start_time + (avg_duration / 2)  # Recalculate for mid-transit if outside window
  }
  
  # Step 6: Prepare the bomb plan output with day of the week, time of explosion, and coordinates
  day_of_week <- wday(bomb_time, label = TRUE)
  return(data.frame(
    Day_of_Week = day_of_week,
    Time_of_Explosion = bomb_time,
    Longitude = bomb_location[1],
    Latitude = bomb_location[2]
  ))
}

# Sample data for testing the algorithm
# Replace these values with real data if testing with specific GPS logs
start_location <- c(-114.0003, 46.88678)  # Example starting location
start_time <- ymd_hms("2024-08-24 17:50:42")  # Example start time
avg_duration <- 9 * 3600  # Average movement duration in seconds (9 hours)
buffer_distance <- 5  # 5 meters for bombing precision
time_buffer <- 10  # 10 seconds for timing precision

# Define stationary and common points for testing purposes
stationary_locations <- data.frame(
  longitude = c(-113.9974, -113.9973),
  latitude = c(46.88527, 46.88590)
)

# Common transit points, replace with actual points if available
common_points <- data.frame(
  longitude = c(-114.0003, -114.0004, -114.0005),
  latitude = c(46.88678, 46.88690, 46.88695)
)

# Generate and print the bomb plan
# This will output the predicted day of the week, time of explosion, and bomb coordinates
bomb_plan <- generate_bomb_plan(start_location, start_time, avg_duration, buffer_distance, time_buffer, stationary_locations, common_points)
print(bomb_plan)

```

